Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-23 sections/eval.tex
Update on Overleaf.

We display speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library, as depicted in Figures \ref{spmv_sorted} and \ref{spmv_grouped}.  We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al. to test the OSKI interface [cite], Ahrens et al. to test a variable block row format partitioning strategy [cite], and Kjolstad et al. to test the TACO library [cite]. Additionally, we included the SNAP graph collection to test with boolean matrices. We also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases. The dense vector is randomly generated. We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset in Figure \ref{spmv_grouped} is the fastest among the formats and programs we tested. Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.

@@

\subsubsection{Finch Logic}
The logical language takes inspiration from relational algebra while incorporating the ordered nature of dimensions in sparse tensors. This means that ev

\begin{align*}
    \finchplan(queries...) &  \quad\quad\quad \finchquery(name, expr) & \finchreorder(expr, idxs...)\\
    \finchrelabel(expr, idxs...)& \quad\quad\quad \finchmapjoin(op, exprs...)  & \finchaggregate(op, expr)\\
    \quad\quad\quad & \quad\quad\quad \finchtable(tns, idxs...)\quad\quad\quad  & \quad\quad\quad \\
\end{align*}

@@

\subsubsection{Lowering}

\subsubsection{Heuristic Optimization}

>>>>

We display speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library, as depicted in Figures \ref{spmv_sorted} and \ref{spmv_grouped}.  We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al. to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al. to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al. to test the TACO library \cite{kjolstad_tensor_2017}. Additionally, we included the SNAP graph collection to test with boolean matrices. We also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases. The dense vector is randomly generated. We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset in Figure \ref{spmv_grouped} is the fastest among the formats and programs we tested. Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.

@@

\subsubsection{Finch Logic}
The expression fragment of the logical language takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor. This means that every operator (with the exception of $\finchtable$ and $\finchalias$ which are analogous to the scan operator in RA) takes in an indexed tensor and outputs an indexed tensor. Conceptually, an indexed tensor can be thought of as a relation with an ordering on the attributes, and we include the $\finchreorder$ operator to manipulate the order of these attributes. This ordering is necessary for two reasons 1) the outputs of a function in the Array API must match a defined order on the dimensions based on the order of the inputs' dimensions 2) to benefit from concordant iteration we must ensure that the operands to a mapjoin have compatible index orders.

To express materialization, reuse of common sub-expressions, and multiple outputs, we define queries and plans. A query is the assignment of the output of an expression to a name. We use this assignment to denote that we materialize an expression in memory. Later queries can then access the result of earlier queries through the alias operator which allows multiple queries to benefit from a shared computation. A plan is a sequence of queries and it produces a set of tensors as output.

\begin{align*}
    \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
    \finchrelabel(expr, idxs...) \quad\quad\quad \finchmapjoin(op, exprs...) \quad\quad\quad \finchaggregate(op, expr)\\
    \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
     expr:= \finchreorder | \finchrelabel |\finchmapjoin | \finchaggregate | \finchtable | \finchalias\quad\quad
\end{align*}

Given this language, we now describe how to define a few example functions from the API as Finch Logic expressions, 
\begin{align*}
\text{sum(M, dims=[2])} &\rightarrow \finchplan(\finchquery(out,\finchaggregate(\finchrelabel(M, i_1,\ldots,i_d), i_2)), out)\\
\text{matmul(A, B)} &\rightarrow \finchplan(\finchquery(out,\finchaggregate(\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)), out)\\
\end{align*}

\subsubsection{Finch Interpreter} 
\textbf{Lowering}

\textbf{Heuristic Optimization}

@@
